---
title: 'Modelling censored log-normal data in `mgcv`'
author: "David L Miller and Michael Dunbar (Environment Agency)"
date: "2026-03-02"
format:
  html:
    math: mathjax
  ipynb:
    math: mathjax
engine: knitr
jupyter: r
execute:
  echo: true
  warning: false
  message: false
  cache: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I show how to model censored log-normally-distributed data. This kind of data often occurs in pollutant modelling where we have some concentration we are measuring which is a positive value, has a potentially skewed distribution, and might be censored at some limit for detection. In that case, the log of the variable tends to look normally distributed. In this case we can model the data as normally distributed, applying the appropriate transformation at the end of the analysis. Unfortunately this implies a [geometric rather than arithmetic mean](https://en.wikipedia.org/wiki/AM%E2%80%93GM_inequality), which can be difficult to compare to other modelling results.

In this notebook, I'll show show to perform this modelling using the popular R package [`mgcv`](https://cran.r-project.org/package=mgcv). You can find out more about generalized additive models [here](https://training.bioss.ac.uk/course_materials/GAMintro2023/). The methods shown here correct for the bias be incur by transforming predictions, ensuring that results are comparable to other modelling that uses the arithmetic mean.


# Modelling preamble


What does it mean to model log-normal data? What we are saying is that we have some variable, Y, which is $\log_{b}$-normally^[Throughout I use $\log_{b}$ to indicate logarithm to the base $b$ since these calculations work regardless of the base used. When the distribution is referred to informally (i.e., non-mathematically) I drop the base.] distributed so $Y\sim\log_{b}N(\mu,\sigma^{2})$ where $\mu$  is the logarithm of the mean and $\sigma^{2}$ is the base $b$ logarithm of the variance of $Y$. This is equivalent to defining a variable $Z=b^{Y}$ where $Z\sim N(\mu_{*},\sigma_{*}^{2})$. The log-normal distribution is useful in cases where $Y$ is defined over the interval $(0,\infty]$ (i.e., $Y>0$).

Many kinds of sensor-collected data will be subject to some "limit of detection" (LoD), where there is some point at which observations are recorded to be zero because the value is smaller than the sensor can detect, and/or the observation is arbitrarily high due to a saturation effect. Censored normals can be used in this case. Core to their formulation is the separation of the observation into censored and non-censored groups, where the censored likelihood contributions are evaluations of the cumulative distribution function (CDF) of the normal distribution. Censored normal distributions are implemented in `mgcv` in the `cnorm` function.


## Arithmetic versus geometric means

We can estimate estimate the expectation of $Y$ ($\mathbb{E}(Y)$) by either an arithmetic or geometric mean (denoted $\mathbb{E}_{A}(Y)$ or $\mathbb{E}_{G}(Y)$, respectively). These are defined as: 

\begin{align}
\mathbb{E}_{A}(Y) &=b^{\mu+\frac{1}{2}\sigma^{2}}\quad\text{and} \\
\mathbb{E}_{G}(Y) &=b^{\mu}.
\end{align}

Where the geometric “mean” corresponds to the median of the distribution. This difference is shown in the following plot

```{r am-gm-fig}
library(ggplot2)

xx <- seq(0, 8, length.out=300)

dat <- data.frame(x = c(xx, xx),
                  dl = c(dlnorm(xx, 0, 1),
                         dlnorm(xx, 1, 1/2)),
                  pars = rep(c("0, 1", "1, 1/2"), c(300,300)))

pardat <- data.frame(pars = rep(c("0, 1", "1, 1/2"), c(2,2)),
                     method = rep(c("Arithmetic", "Geometric"),
                                  2),
                     value = exp(c(0+1/2*1^2, 0,
                               1+1/2*1/2^2, 1)))

ggplot(dat) +
  geom_line(aes(x=x, y=dl, group=pars, colour=pars)) +
  geom_vline(aes(xintercept=value, lty=method, group=pars,
                 colour=pars), data=pardat) +
  labs(x="x", y="Probability density", colour="Parameters",
       lty="Estimator") +
  theme_minimal()
```

Similarly, we define the arithmetic and geometric standard deviations as follows:

\begin{align}
\text{SD}_{A}(Y) &=b^{\mu+\frac{1}{2}\sigma^{2}}\sqrt{b^{\sigma^{2}}-1}\\
\text{SD}_{G}(Y) &=b^{\sigma}.
\end{align}

This is all fine in the simple case where we're just estimating the distribution parameters, with no covariates involved. What about when we want to use covariates for prediction?

## log-normal response in regression-like models

We may want to model the mean via some linear model, for example when using a GLM or GAM. We decompose the mean as $\mathbf{X}\boldsymbol{\beta}$, where $\mathbf{X}$ is an $n\times p$ design matrix and $\boldsymbol{\beta}$ is a $p$-vector of coefficients to be estimated (where n is the number of observations).

In practice, one can simply assume that the log-response is normally distributed and fit a GLM-a-like (GAM/GLMM/GAMM/etc) with a normal distribution. This model is:

\begin{equation*}
\mathbb{E}(Y)=b^{\mu}=b^{\mathbf{X}\boldsymbol{\beta}},
\end{equation*}

which is clearly modelling the geometric, rather than arithmetic mean (equivalently, we are modelling the median). This naturally extends to the case of random effects and therefore penalized smoothers (i.e., GLMMs, GAMs and GAMMs).

We can post hoc correct predictions to use the arithmetic mean using a multivariate version of the corrections above. These corrections lead to the following expressions:

\begin{align*}
\mathbb{E} \left[ g \left( \hat{\boldsymbol{\beta}} \right) \right] &= g\left( \hat{\boldsymbol{\beta}} \right) + \frac{1}{2n}\text{trace}\left(D^{2}g(\hat{\boldsymbol{\beta}})\Sigma\right)\quad\text{and}\\
\hat{\text{Var}}\left[g\left(\hat{\boldsymbol{\beta}}\right)\right] & = \frac{1}{n}\left[Dg(\boldsymbol{\beta})\right]{}^{T}\Sigma(Dg(\boldsymbol{\beta})) -\frac{1}{2n^{2}}\text{trace}\left(D^{3}g(\boldsymbol{\beta})D^{2}g(\boldsymbol{\beta})\Sigma^{2}-\left[D^{2}g(\boldsymbol{\beta})\Sigma\right]^{2}\right)
\end{align*}

where $g(\boldsymbol{\beta})=b^{X\boldsymbol{\beta}}$ if $b$ is the base we are working in and $X$ is the design matrix for our model. $D$ indicates the derivative operator with respect to $\boldsymbol{\beta}$, with powers of $D$ giving that order of derivative. $\Sigma$  is the posterior covariance matrix for $\boldsymbol{\beta}$ (see, e.g. [Miller (2025)](https://doi.org/10.1111/2041-210X.14498) for further explanation of covariance matrices in the context of empirical Bayes modelling).




# Analysis of pollutant data

We now start our analysis. This code is based on examples by Mike Dunbar to illustrate fitting a GAMM with censored normal family to water quality time series from multiple sites ([Harmonised Monitoring Scheme sites for England](https://en.wikipedia.org/wiki/Harmonised_monitoring_scheme)) to look at smoothed trends over time and the effects of seasonality. For each determinand we fit a separate model where site is a random effect and we use two temporal smoothers.

We begin by loading the requisite packages.

```{r libs}
library(dplyr)
library(ggplot2)
library(mgcv)
library(tictoc)
library(lubridate)
library(tidyr)
library(purrr)
library(gratia)
library(stringr)
```

There are some non-CRAN packages required which we can install directly from GitHub, then load:

```{r extra-libs}
# for nice scales
remotes::install_github("christyray/sciscales")
library(sciscales)

# for the censored log normal helper functions
remotes::install_github("dill/mgcvUtils")
library(mgcvUtils)
```

Michael Dunbar at the Environment Agency pre-prepared data for this analysis using open data from EA (extracted using EA APIs).

```{r data}
# created in data_prep.R
load('data/HMS_open_data.rda')
head(samples)
```

This data has four determinands, for brevity here we'll just look at one: Nitrate (where `determinand.notation==0117` in our data).

```{r nitrate-only}
samples <- samples |>
  filter(determinand.notation == "0117")
```

## Data checking

Plotting the data, we can see how the limit of detection (red) changes over time.^[The censored data is coded in the `resultQualifier.notation` columns with a `<`.]

```{r data-check}
samples |>
  ggplot() +
  geom_point(aes(x = sample.sampleDateTime,
                 y = result,
                 group = sample.samplingPoint.notation,
                 colour = resultQualifier.notation)) +
  scale_y_log10() +
  labs(x="Date", y="Nitrate", colour="LoD") +
  theme_bw()
```

```{r}

# tabulate LoDs
samples |>
  # can comment out
  filter(resultQualifier.notation == '<') |>
  group_by(result) |>
  summarise(n = n())
```


The predominant LoD in the original data is 0.196 although some earlier data it is 0.5.

## Final data prep

When we log-transform we can't use zero as the lower bound for censored values
we just choose a very small value. We use a value 2 orders of magnitude below lowest limit of detection.

```{r zeros}
nearly.zero <- 0.000001
log.zero <- log10(nearly.zero)
```

Create some extra columns, ensure categorical variables are factors.

```{r}
samples <-
  samples |>
  mutate(DATE = date(sample.sampleDateTime),
         year = year(DATE),
         month = month(DATE),
         samplingPoint.F = as.factor(sample.samplingPoint.notation),
         det.code.desc = paste(determinand.notation, determinand.label, sep = '-'),
         DATE_DEC = round(decimal_date(DATE),2),
         MEAS_SIGN = replace_na(resultQualifier.notation, '='),
         log.meas.upper = log10(result + nearly.zero),
         log.meas.lower = if_else(MEAS_SIGN == '<', log.zero, log.meas.upper),
         censored = ifelse(MEAS_SIGN == '<', TRUE, FALSE),
         # these two lines are because for the not artificially censored data,
         # the ..original columns won't be populated yet
         result.original = result,
         qualifier.original = resultQualifier.notation
         )
```

Finally we create the response matrix, which is two columns (if entries are the same there was no censoring, if they are different there was censoring; see `?mgcv::cnorm`):

```{r response}
samples$y <- cbind(samples$log.meas.lower, samples$log.meas.upper)
head(samples$y)
```

# Modelling

We can now fit our generalized additive model. We want to fit a model of the following form:

\begin{equation*}
\mathbb{E}(Y_i) = b^{\beta_0 + s(t_i) + s(m_i) + \beta_{z_i}}
\end{equation*}

where $i$ indexes the observations which have nitrate level $Y_i$ on date $t_i$ in month $m_i$ and station $z_i$ (where $\beta_{z}$ is a random intercept for station $z$). Generic smoothers are denoted $s()$. We want to look at any common trend across our monitoring sites, which will be in the terms $s(t_i)$ and $s(m_i)$. We use $b=10$ for this modelling, though any scientifically appropriate base would be usable with these functions/methods.

Here we will use [`bam` and fREML](https://doi.org/10.1080/01621459.2016.1195744) for speed here to show that these methods work with large data approaches too.

We need to setup knots for the monthly smoother. Note that we just need to specify the first and last knot, then `mgcv` will fill in evenly-spaced knots in that range. See [this useful blog post by Gavin Simpson about cyclic smoothing](https://fromthebottomoftheheap.net/2014/05/09/modelling-seasonal-data-with-gam/).

```{r cyclic-knots}
# for month cyclic cubic spline
sknots <- list(month=c(0.5, 12.5))
```

We can now fit the model

```{r modelling}
b <- bam(y ~ s(DATE_DEC, bs = 'cr') +
             s(month, bs = 'cc') +
             s(samplingPoint.F, bs = 're'),
         method = 'fREML',
         family = clognorm(base=10),
         knots = sknots,
         data = samples)
```

Looking at a summary output as usual. In particular we can see that we used the new `clognorm` family (see `Family:` line). This shows the base (10) and the standard deviation result.

```{r model-summary}
summary(b)
```

# Making predictions


We want to make some predictions over the date range of the data (no extrapolation). We are interested in patterns with and without the seasonality (month) component.

```{r predictions, eval=FALSE}
date.sequence <- expand_grid(year = 2000:2025, month = 1:12) |>
  mutate(DATE_DEC = decimal_date(ymd(paste(year, month, '15', sep = '-')))) |>
  # don't want to predict past September 2025 (even so, later data could be
  # patchy)
  filter(!(year == 2025 & month %in% c(10, 11,12))) |>
  as.data.frame()


# this is a version of our date sequence with an additional column for site
# this is necessary because our model has site in it so the prediction matrix
# needs it
# !! this is the case even if we are going to exclude site from the predictions
ds.1site <-
  date.sequence |>
  mutate(samplingPoint.F = samples.2$samplingPoint.F[1])
```



```{r what, eval=FALSE}
# HELPER FUNCTIONS
post_process_predictions <- function(pred.df) {
  cols1 = c(gm.conc = '.fitted',
            gm.lower.ci = '.lower_ci',
            gm.upper.ci = '.upper_ci')

  pp <-
    pred.df |>
    mutate(.var = .se ^2,
           across(all_of(cols1), ~if_else(transform == 'log', 10^.x, .x)),
           Date = ifelse(is.na(year), NA,
                         decimal_date(ymd(paste(year, month, '15', sep='-')))),
           season = factor(case_match(month,
                                      3:5 ~ 'Spring',
                                      6:8 ~ 'Summer',
                                      9:11 ~ 'Autumn',
                                      c(12,1,2) ~ 'Winter'),
                           levels = c('Spring', 'Summer', 'Autumn', 'Winter')))

  return(pp)
}

# note that it is OK to do exclude = '' or exclude = NULL, the exclude is then
# just ignored
build_exclude <- function(include_site_flag=FALSE, include_season_flag=TRUE){
  out <- character(0)
  if (!include_season_flag) out <- c(out, "s(month)")
  if (!include_site_flag)  out <- c(out, "s(samplingPoint.F)")
  out
}


predict_and_collate <-
  function(my.data, my.model.list, include.site.flag = FALSE) {

    cat('Making and collating predictions\n')

    exclude.noseas <- build_exclude(include.site.flag,
                                    include_season_flag = FALSE)
    exclude.withseas <- build_exclude(include.site.flag,
                                      include_season_flag = TRUE)

    fv.list <- list()

    cat('overall mean without seasonality\n')
    fv.list[['Without season']] <-
      fitted_values(my.model.list$bam.mod.a,
                    data = my.data ,
                    scale = "response",
                    exclude = exclude.noseas)

    cat('overall mean with seasonality\n')
    fv.list[['With season']] <-
      fitted_values(my.model.list$bam.mod.a,
                    data = my.data,
                    scale = "response",
                    exclude = exclude.withseas)

    fv <- fv.list |>
      list_rbind(names_to = 'Prediction type') |>
      # this column not used here: it's used in other cases where we're fitting
      # some models to raw data and some to log data
      mutate(transform = 'log') |>
      post_process_predictions()

    return(fv)

  }

# END OF HELPER FUNCTIONS

# make predictions
# separate loop for simplicity

# we create a receptable for the predictions for each det
super.fv.list <- list() # this always has site excluded
observed.fv.list <- list()

for(i in names(model.list.b)) {
  for(j in my.dets$det.code.desc) {

    cat('-----------------------------------\n')
    cat('Doing model: ', i, ' Doing det: ', j, "\n")

    super.fv.list[[i]][[j]] <-
      # not strictly necessary to pass a list - this functionality left in in
      # case we have more than one model formulation for each det
      predict_and_collate(
        ds.1site,
        list(bam.mod.a = model.list.b[[i]][[j]]), include.site.flag = FALSE)

    cat('predictions for observed data\n')
    # this is a bit of a fudge: we pass the observed data to predict_and_collate
    # we should note that when comparing with observed its the predictions with
    # seasonality that are most relevant
    # also here we are predicting for EVERY SITE so the output will be much
    # bigger
    observed.fv.list[[i]][[j]] <-
      predict_and_collate(
        samples.2 |> filter(det.code.desc == j),
        list(bam.mod.a = model.list.b[[i]][[j]]), include.site.flag = TRUE)

  }
}

# end up finally with one big data frame of predictions
fv.final <-
  super.fv.list |>
  map(list_rbind, names_to = 'det.code.desc') |>
  list_rbind(names_to = 'model.type') |>
  # remove predictions for older PFOS data = don't trust the raw data from that
  # period
  filter(!(det.code.desc == '9276-PFOS' & year <= 2016 ))

# reminder check what combinations we have
fv.final |>
  select(model.type, det.code.desc, `Prediction type`) |>
  distinct()

fv.observed <-
  observed.fv.list |>
  map(list_rbind, names_to = 'det.code.desc') |>
  list_rbind(names_to = 'model.type') |>
  filter(`Prediction type`== 'With season') |>
  select(model.type, samplingPoint.F, det.code.desc, DATE_DEC, MEAS_SIGN,
         log.meas.upper, log.meas.lower, gm.conc, gm.lower.ci, gm.upper.ci,
         resultQualifier.notation, result, qualifier.original,
         result.original) |>
  mutate(residual = result.original - gm.conc)

fv.observed |>
  select(model.type, det.code.desc) |>
  distinct()


# 1. PLOT PREDICTIONS WITH fv.final - averaged across sites, regular time step,
# with and without seasonality
# we want to compare the 2 graphs in the top row and the 2 graphs in the  left
# column
# No need to compare the other combinations

fv.final |>
  # comment out if using facet_grid to show a 2x2 plot of dets and model/data
  # types
  # remember that with the original data, even when we substitute, there so
  # little censoring it's basically the same dataset
  #filter(model.type == 'cen') |>
ggplot() +
  # un-comment line to include raw data: <LoDs are plotted at the LoD (plot
  # takes alot longer to render with the raw data)
  #geom_point(data = samples.2, aes(x = DATE_DEC, y = result), col = 'grey',
  #           alpha = 0.5, pch = '+') +
  geom_line(aes(x = DATE_DEC, y = gm.conc, col = `Prediction type`))  +
  geom_ribbon(aes(x = DATE_DEC, ymin = gm.lower.ci, ymax = gm.upper.ci, fill = `Prediction type`),
              alpha = 0.2) +
  #facet_wrap(~det.code.desc, ncol = 1, scales = 'free_y') +
  facet_grid(rows = vars(model.type), cols = vars(det.code.desc)) +
#  scale_y_log10(labels = label_sci()) +
  scale_y_log10() +
  scale_color_brewer(palette = 'Dark2') +
  scale_fill_brewer(palette = 'Dark2', guide = 'none') +
  labs(x = 'Date', y = 'Concentration', caption = 'Concentration units are mg/l for nutrients and (\u00b5g/l) for metals and chemicals') +
  theme_bw() +
  theme(legend.position = 'bottom')


# PLOTTING original OBSERVED against the predictions for the censored model
# remember the artificial censoring was applied at 2 mg/l but we retained the
# original values as result.original
# obviously the model won't be able to reproduce the range of the original data
# and also this is a random intercept model, not a factor smooth model
# fs will almost certainly be better but I didn't have time / computing power
# to fit it
# we're looking for potential bias both above and below the original detection
# limit
fv.observed |>
  filter(model.type == 'cen') |>  # & det.code.desc == "0117 artificially censored-Nitrate-N") |>
  # there are a minority of very low observed readings, I don't think we need
  # to worry about these in this case
  filter(result.original > 0.1) |>
  # retain every xth row to speed up plotting
  filter(row_number() %% 5 == 1) |>
  ggplot() +
  geom_point(aes(x = result.original, y = gm.conc), col = 'grey', pch = '+') +
  geom_abline(aes(intercept = 0, slope = 1)) +
  # original censoring
  geom_vline(aes(xintercept = 2), col = 'red') +
  facet_wrap(~det.code.desc) +
  scale_x_log10() + scale_y_log10() +
  coord_fixed(ratio = 1) +
  theme_bw()

# attempt to plot 'residuals'
# conclusion - we are seeing some general pattern in residuals particularly
# below the artificial LoD
fv.observed |>
  filter(model.type == 'cen' &
         det.code.desc == "0117 artificially censored-Nitrate-N") |>
  # there are a minority of very low observed readings, I don't think we need
  # to worry about these in this case
  #filter(result.original > 0.1) |>
  filter(row_number() %% 5 == 1) |>
  # these are 'residuals' in inverted commas because we're comparing original
  # (virtually uncensored) values with
  # predictions from a model where artificial censoring was introduced
  mutate(residual = result.original - gm.conc) |>
  ggplot() +
  geom_point(aes(x = gm.conc, y = residual), col = 'grey', pch = '+') +
  geom_vline(aes(xintercept = 2), col = 'red') +
  theme_bw()
```


